\message{ !name(mainfile.tex)}\documentclass[a4,danish]{article}
\input{preamble}
\input{frontpage}

\begin{document}

\message{ !name(L2-metric.tex) !offset(-7) }
\subsection{The L2 metric vanishes}
\label{sec:l2-metric-vanishes}

In the previous section we went through some work to construct a measure of length for a path $q$ in the space $\mathcal{I}$. As the tangent spaces were seen to consist of some functions on the circle, it was natural to consider using a version of the L$^2$-metric (a version invariant to reparametrizations). However, as we illustrate in this section, this does not give rise to a usable notation of length, because the geodesic distance following from this metric becomes 0 for every two curves in the space.

It can seem weird to go through so much trouble to define a useless distance. But firstly, the construction illustrates some of the difficulties in defining a reasonable notion on length on $\mathcal{I}$; secondly, and more importantly, the fact that the distance vanishes is a quite surprising result, which depends crucially on the formula for the length of a path given in Proposition~\ref{prop:length-quotient}. This gives us some justification for the somewhat unfruitfull work of definiting the L$^2$-metric.

The ``proof'' below is mostly heuristic, and we only consider the very simply case of a path transforming the circle into a larger circle, but this captures the main idea.
\begin{theorem}
  \label{theorem:l2-metric-vanishes}
  Let $a,b \in \mathcal{I}$. Then for every $\epsilon >0 $ there exists a path $q$ in $\mathcal{I}$ such that
  \begin{equation*}
    \mathcal{L}(q) < \epsilon.
  \end{equation*}
\end{theorem}
\begin{corollary}
  For all $a,b \in \mathcal{I}$
  \begin{equation*}
    \mathcal{D}(a,b) = 0.
  \end{equation*}
\end{corollary}

\begin{proof}[Proof of Theorem~\ref{theorem:l2-metric-vanishes}]
  Start by rewriting the \eqref{eq:length-quotient} from Proposition~\ref{prop:length-quotient} as follows.
  \begin{align*}
    \mathcal{L}(q) &
                     = \int_{0}^{1}
                     \left(
                     \int_{\S^{1}} \frac{\langle q_{t},i q_{\theta}\rangle^2}
                     {|q_{\theta}|}  \diff
                     \right)^{1/2} \theta \diff t \\
                   & =  \int_{0}^{1}
                     \left(
                     \int_{\S^{1}}
                     \left(
                     \frac{\langle q_{t},i
                     q_{\theta}\rangle}{| q_{t}|| q_{\theta}|}
                     \right)^2
                     | q_t|^2   | q_{\theta}|
                     \diff \theta
                     \right)^{1/2} \diff t \\
                   &  =
                     \int_{0}^{1}
                     \left(
                     \int_{\S^{1}}
                     \cos(\alpha( q_t, i q_{\theta}))^2
                     | q_t|^2   | q_{\theta}|
                     \diff \theta
                     \right)^{1/2}\diff t,
  \end{align*}
with $\alpha(x,y)$ denoting the angle between $x$ and $y$. When
constructing a zigzag-path the angle will be \hl{(approximately?)}
constant in $\theta$ and $t$,
and it is given by
\begin{equation*}
  n = \tan(\alpha).
\end{equation*}
\end{proof}

We have that
\begin{equation*}
  \cos(\arctan(n)) = (1+n^2)^{-1/2}
  = O(n^{-1}),
\end{equation*}
so we can write
\begin{equation*}
  L(\phi) = O(n^{-2})
  \int_{0}^{1} \int_{\S^{1}}
  |\phi_t|^2   |\phi_{\theta}|
  \diff \theta \diff t,
\end{equation*}
in this case. To show that such a zigzag path has arbitrary
small length we just need to show that the integral does not grow
faster than $n^2$.

As an example, take the simply case where we expand the circle $e^{i\pi\theta}$ to
$2e^{i\pi\theta}$. The zigzag path in then concretely given as
\begin{equation*}
  \phi(t,\theta) = e^{i\pi\theta}
    \sum_{k=0}^{n-1}
    h^{n,k}(t,\theta) + g^{n,k}(t,\theta)
\end{equation*}
where
\begin{equation*}
  \begin{aligned}
    h^{n,k}(t,\theta) & := 1_{[\frac{k}{n},\frac{k}{n} +
      \frac{1}{2n})}(\theta) \left( 1+2t(n\theta-k) \right), \\
    g^{n,k}(t,\theta) & := 1_{[\frac{k}{n} + \frac{1}{2n},\frac{k+1}{n})}(\theta)
    \left( 1+2t(1-n\theta-k) \right).
  \end{aligned}
\end{equation*}

We have that
\begin{equation*}
  \begin{aligned}
    |\phi_t| & = \sum_{k=0}^{n-1} |h^{n,k}_t| + \sum_{k=0}^{n-1}
    |g^{n,k}_t|, \\
    |\phi_{\theta}| & = \sum_{k=0}^{n-1} |h^{n,k}_{\theta} + h^{n,k} | +
    \sum_{k=0}^{n-1}
    |g^{n,k}_{\theta} + g^{n,k} |,
  \end{aligned}
\end{equation*}
so by symmetry
\begin{equation*}
  \begin{aligned}
    \int_{0}^{1}
    |\phi_t|^2   |\phi_{\theta}|
    \diff \theta
    & =
    2n \int_{0}^{\frac{1}{2n}} |h^{n,0}_t|^2 |h^{n,0}_{\theta} + h^{n,0} |
    \diff \theta \\
    & = 2n \int_{0}^{\frac{1}{2n}}
    (2n\theta)^2(2tn+1+t2n\theta) \diff \theta \\
    & = \int_{0}^1
    u^2(2tn+1+t u) \diff \theta \\
    & = O(n),
  \end{aligned}
\end{equation*}
for $t\in[0,1]$ which gives the result.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "mainfile"
%%% End:

\message{ !name(mainfile.tex) !offset(-125) }

\end{document}
